Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-04 ms/ms.Rnw
updated phylo methods section

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species were classified as shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}. [NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]

@@

I excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo}). The R code accompanying this paper documents these decisions with citations to the relevant literature.

@@

I tested whether Raunki\ae r life form was associated with \el~among British angiosperms using ANOVA with Type-2 sum of squares. I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. See the R code accompanying this paper for further detail. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.

@@

I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo}
\end{figure}

@@

>>>>

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species are shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}. I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).

@@

I excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees). I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail). Specifically, the estimated $\alpha$ parameter was extremely high. In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e. phylogenetic signal). When there is no phylogenetic signal (i.e. high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates. Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}. Therefore, I used a regular ANOVA with Type-2 sum of squares.

@@

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo-lifeform}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_gf-light.pdf}}
\caption{Growth forms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different growth forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:gf-light}
\end{figure}

### 2017-10-04 ms/ms.tex
added figure comparing growth forms
modified table 1
updated phylo methods section

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species were classified as shrubs (plant height less than 4 m) or trees (plant height greater than 4 m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}. [NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]

@@

I excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (3 species) and CAM (2 species) plants. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 372 species (Fig.~\ref{fig:phylo}). The R code accompanying this paper documents these decisions with citations to the relevant literature.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.2
  & 90.4
  & 0 \\
    \el~+ lifeform
  & 0.46
  & 0.071
  & 0.32
  & 8
  & -40.2
  & 96.4
  & 6 \\
    \el
  & 0.64
  & 0.107
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.2 \\
    lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 82 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.6
  & 221.1
  & 130.7 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

>>>>

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species are shrubs (plant height less than 4 m) or trees (plant height greater than 4 m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}. I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).

@@

I excluded C$_4$ (3 species) and CAM (2 species) plants. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.2
  & 90.4
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.4
  & 9.9 \\
    \el~+ Raunki\ae r lifeform
  & 0.46
  & 0.071
  & 0.32
  & 8
  & -40.2
  & 96.4
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.4
  & 102.7
  & 12.3 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 82 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 179.1
  & 88.6 \\
    \el
  & 0.64
  & 0.107
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.2 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.6
  & 221.1
  & 130.7 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{CAPTION}
\label{fig:compare-growth-forms}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_gf-light.pdf}}
\caption{Growth forms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different growth forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:gf-light}
\end{figure}
